:::{#cor-mom-vc}
#### Method-of-moments estimators of the variance components

Under @def-one-way-re,
$\est\sigma^2 \eqdef \text{MSE}$ and
$$
\est\sigma_\alpha^2 \eqdef \frac{\text{MSA} - \text{MSE}}{m}
$$ {#eq-mom-vc}
are unbiased estimators of $\sigma^2$ and $\sigma_\alpha^2$.
:::

{{< slidebreak >}}

::: proof

This proof uses @thm-ems-one-way.

$$
\ba
\E{\est\sigma_\alpha^2}
&= \E{\frac{\text{MSA} - \text{MSE}}{m}}
&& \text{(definition of the estimator)}
\\
&= \frac{\E{\text{MSA}} - \E{\text{MSE}}}{m}
&& \text{(linearity of expectation)}
\\
&= \frac{m\sigma_\alpha^2 + \sigma^2 - \sigma^2}{m}
&& \text{(expected mean squares)}
\\
&= \frac{m\sigma_\alpha^2}{m}
&& \text{(cancel } \sigma^2\text{)}
\\
&= \sigma_\alpha^2
&& \text{(cancel } m\text{)}
\ea
$$ {#eq-mom-vc-unbiased}

$\E{\est\sigma^2} = \E{\text{MSE}} = \sigma^2$ directly from @thm-ems-one-way.
:::

{{< slidebreak >}}

:::{#exm-dyestuff-vc}
#### Variance components for the dyestuff batches

The ANOVA table for @exm-dyestuff gives the mean squares:

```{r}
#| label: dyestuff-anova
#| code-fold: show
anova(lm(Yield ~ Batch, data = Dyestuff))
```

By @cor-mom-vc,
$\est\sigma^2 = 2451.3$ and
$$
\ba
\est\sigma_\alpha^2
&= \frac{11271.5 - 2451.3}{5}
&& \text{(substitute the mean squares and } m = 5\text{)}
\\
&= \frac{8820.2}{5}
&& \text{(subtract)}
\\
&= 1764.04
&& \text{(divide)}
\\
&\approx 1764.0
&& \text{(round to one decimal place)}
\ea
$$ {#eq-dyestuff-mom-vc}

The `lmer()` function in the `lme4` package [@bates2015lme4]
fits the same model by restricted maximum likelihood (REML);
`(1 | Batch)` requests a random intercept for each batch:

```{r}
#| label: dyestuff-lmer
#| code-fold: show
dyestuff_lmer <- lmer(Yield ~ 1 + (1 | Batch), data = Dyestuff)
VarCorr(dyestuff_lmer) |> as.data.frame()
```

In this balanced design with a positive moment estimate,
the REML estimates match the method-of-moments estimates.
:::

The method-of-moments estimate $\est\sigma_\alpha^2$ is negative whenever $\text{MSA} < \text{MSE}$,
which is impossible for a variance;
likelihood-based methods such as REML constrain the estimate to be nonnegative,
and they also handle unbalanced designs
[@rocke2021epi204; @bates2015lme4].
