:::{#exm-radon-shrinkage}
#### Shrinkage of county mean radon levels

We fit the random-intercept model to the radon data from @exm-radon-data:

```{r}
#| label: radon-lmer0
#| code-fold: show
radon_lmer0 <- lmer(log_radon ~ 1 + (1 | county), data = radon)
VarCorr(radon_lmer0) |> as.data.frame()
fixef(radon_lmer0)
```

The estimates are
$\est\mu \approx 1.350$,
$\est\sigma_\alpha^2 \approx 0.0886$,
and $\est\sigma^2 \approx 0.587$.
Aitkin County has $n_j = 4$ houses with sample mean $\bar{y}_j \approx 0.715$.
Plugging these estimates into @thm-partial-pooling:

$$
\ba
w_j
&= \frac{4 / 0.587}{4 / 0.587 + 1 / 0.0886}
&& \text{(substitute the estimates)}
\\
&\approx \frac{6.81}{6.81 + 11.29}
&& \text{(divide)}
\\
&\approx 0.377
&& \text{(divide)}
\\
\est\mu_j
&\approx 0.377 \times 0.715 + 0.623 \times 1.350
&& \text{(substitute into the weighted average)}
\\
&\approx 1.111
&& \text{(multiply and add)}
\ea
$$

Aitkin County's estimate moves most of the way from its own mean toward the overall mean.
In contrast, Hennepin County has $n_j = 105$ houses,
so its weight is about 0.94 and its estimate barely moves.
The same calculation for every county
reproduces the estimates that `lme4` reports:

```{r}
#| label: radon-shrinkage-check
#| code-fold: show
radon_vc <- as.data.frame(VarCorr(radon_lmer0))$vcov
radon_w <- (radon_n / radon_vc[2]) /
  (radon_n / radon_vc[2] + 1 / radon_vc[1])
radon_ybar <- tapply(radon$log_radon, radon$county, mean)
radon_pp <- radon_w * radon_ybar + (1 - radon_w) * fixef(radon_lmer0)
max(abs(radon_pp - coef(radon_lmer0)$county[, 1]))
```
:::

{{< slidebreak >}}

:::{#fig-radon-shrinkage}

```{r}
#| label: radon-shrinkage-plot
#| code-fold: true
plot(
  as.vector(radon_n), radon_ybar,
  log = "x", pch = 1,
  xlab = "Houses sampled in county (log scale)",
  ylab = "Estimated county mean log radon"
)
points(as.vector(radon_n), radon_pp, pch = 19)
abline(h = fixef(radon_lmer0), lty = 2)
legend(
  "topright",
  legend = c("No pooling", "Partial pooling"),
  pch = c(1, 19)
)
```

No-pooling and partial-pooling estimates of the county mean log radon levels;
the dashed line is the complete-pooling estimate.
:::
