Covariates enter a multilevel model as fixed effects,
and their coefficients can also vary by group
[@gelman2006data, Chapter 12].
For house $i$ in county $j$ with floor indicator $x_{ij}$:

- the **varying-intercept** model is
$Y_{ij} = \beta_0 + \alpha_j + \beta_1 x_{ij} + \epsilon_{ij}$;
- the **varying-intercept, varying-slope** model is
$Y_{ij} = \beta_0 + \alpha_j + (\beta_1 + \gamma_j) x_{ij} + \epsilon_{ij}$,
where each county's pair $(\alpha_j, \gamma_j)$ is bivariate normal with mean zero.

{{< slidebreak >}}

:::{#exm-radon-floor}
#### Radon by floor of measurement

```{r}
#| label: radon-floor
#| code-fold: show
radon_lmer1 <- lmer(log_radon ~ floor + (1 | county), data = radon)
radon_lmer2 <- lmer(log_radon ~ floor + (1 + floor | county), data = radon)
fixef(radon_lmer1)
VarCorr(radon_lmer2)
```

In the varying-intercept model,
measurements on the first floor average about 0.66 lower on the log scale than basement measurements in the same county,
a ratio of $\exp(-0.66) \approx 0.52$.
The varying-slope model lets that difference vary by county.

To compare the two models, `anova()` refits both by maximum likelihood
and computes a likelihood ratio test:

```{r}
#| label: radon-floor-lrt
#| code-fold: show
anova(radon_lmer1, radon_lmer2)
```

The varying-slope model adds two parameters
(the slope variance and its correlation with the intercept),
and the likelihood ratio test gives $p \approx 0.26$.
Because the null value of the slope variance, zero, is on the boundary of the parameter space,
this $p$-value is conservative
[@rocke2021epi204];
AIC and BIC also favor the simpler model.
:::

{{< slidebreak >}}

:::{#exm-radon-uranium}
#### A county-level predictor

Soil uranium is measured once per county,
so it is a **group-level predictor**.
Adding it to the varying-intercept model:

```{r}
#| label: radon-uranium
#| code-fold: show
radon_lmer3 <- lmer(
  log_radon ~ floor + log_uranium + (1 | county),
  data = radon
)
fixef(radon_lmer3)
VarCorr(radon_lmer3) |> as.data.frame()
```

The estimated between-county variance falls from about 0.099 without uranium
to about 0.022 with it,
while the within-county variance barely changes:
county uranium explains most of the differences between counties.
:::
