A model can have more than one level of random effects.
When each lower-level group belongs to exactly one higher-level group,
the random effects are **nested**
[@rocke2021epi204].

{{< slidebreak >}}

:::{#def-nested-re}
#### Two-level nested random-effects model

Suppose $\ell$ laboratories each analyze $b$ batches,
with $r$ replicate measurements per batch.
The **two-level nested random-effects model** is

$$
Y_{ijk} \eqdef \mu + L_i + B_{ij} + \epsilon_{ijk},
$$ {#eq-nested-re}

where $L_i \siid \ndist{0, \sigma_L^2}$ is the effect of laboratory $i$,
$B_{ij} \siid \ndist{0, \sigma_B^2}$ is the effect of batch $j$ within laboratory $i$,
$\epsilon_{ijk} \siid \ndist{0, \sigma^2}$ is the replicate error,
and all three sets of random variables are mutually independent.
:::

By the same argument as @thm-ems-one-way,
the expected mean squares of the nested ANOVA are
$\E{\text{MS}_L} = br\sigma_L^2 + r\sigma_B^2 + \sigma^2$,
$\E{\text{MS}_B} = r\sigma_B^2 + \sigma^2$,
and $\E{\text{MS}_E} = \sigma^2$
[@rocke2021epi204].
So the correct $F$ test for laboratories divides $\text{MS}_L$ by $\text{MS}_B$,
not by $\text{MS}_E$.

{{< slidebreak >}}

:::{#exm-coop}
#### A cooperative trial

The `coop` data in the `MASS` package [@venables2002modern]
come from a cooperative trial
in which $\ell = 6$ laboratories each analyzed $b = 3$ batches,
a month apart,
with $r = 2$ replicates per batch.
We use the measurements of specimen `S1`:

```{r}
#| label: coop-anova
#| code-fold: show
coop2 <- subset(MASS::coop, Spc == "S1")
anova(lm(Conc ~ Lab + Lab:Bat, data = coop2))
```

The `Lab:Bat` term is the batch-within-laboratory effect.
The $F$ statistic that `anova()` prints for `Lab` uses $\text{MS}_E$ as its denominator,
which is wrong here;
the correct statistic is
$F = 0.37804 / 0.01703 \approx 22.2$ on $(5, 12)$ degrees of freedom.

The equivalent mixed model has a random intercept for each laboratory
and for each batch within each laboratory:

```{r}
#| label: coop-lmer
#| code-fold: show
coop_lmer <- lmer(Conc ~ 1 + (1 | Lab) + (1 | Lab:Bat), data = coop2)
VarCorr(coop_lmer) |> as.data.frame()
```

The estimated variance components are
$\est\sigma^2 \approx 0.0063$ (same-day replicates),
$\est\sigma_B^2 \approx 0.0054$ (batches a month apart),
and $\est\sigma_L^2 \approx 0.0602$ (laboratories).
Most of the variation in the measurements comes from differences between laboratories.
:::

Testing whether a variance component is zero is harder than testing a fixed effect,
because $\sigma^2 = 0$ lies on the boundary of the parameter space:
a variance cannot be negative,
so the usual large-sample distribution of the likelihood ratio statistic does not apply
[@rocke2021epi204].
